Robust Control of a Fluxonium Qubit

### 2020-10-12 paper/main.tex
main.tex

($T_{1}$ and $T_{\phi}$ noise)
In addition to constructing gates
that obey experimental constraints and have high simulated fidelities,
we also desire to make the gates robust to noise that affects the experimental
gate fidelity. Decoherence of the quantum state due to external noise
is typically modeled by two phenomena: longitudinal relaxation and pure dephasing.
They are modeled using their $1/e$ decay times $T_{1}$ and $T_{\phi}$ respectively
(see Appendix).
The main contributions to longitudinal relaxation in our
device are dielectric loss in the capacitor, resistive loss in the inductor,
and Purcell loss. The main contributions to pure dephasing in our
device are $1/f$ flux noise and decay via charge and flux coupling
to the control lines.

Dissipation to the thermal bath via longitudinal
relaxation is an irreversible process
that results in information loss.
Converesely, pure dephasing is a reversible process.
There is a tradeoff between the two decoherence processes. In the case of white
noise we have that the sum of the noise weights $W_{1}$ and $W_{\phi}$
is constant \cite{huang2020engineering}.
Our device achieves its best pure dephasing
protection at the flux frustration point
$T_{2e}(a = 0) \sim 300 \mu\textrm{s}$
where the qubit is first-order insensitive to changes in flux.
It becomes more succeptable to pure dephasing as the flux is tuned away from the flux
frustration point $|a| > 0$. Conversely, longitudinal relaxation is at a minimum
at the flux frustration point $T_{1}(a = 0) = 0.315$ms,
and increases away from the flux frustration point
$T_{1}(a = 0.34 \textrm{GHz}) = 4.3$ms. Given the nature
of the decay processes and the tradeoff, we choose
to maximize the longitudinal relaxation time directly
via the flux amplitude and employ robust control techniques to mitigate
pure dephasing.

%% F1
\begin{figure*}[ht]
  \begin{subfigure}{.315\textwidth}
    \includegraphics[width=\linewidth]{assets/f1a.png}
  \end{subfigure}\hfill
  \begin{subfigure}{.23\textwidth}
    \includegraphics[width=\linewidth]{assets/f1b.png}
  \end{subfigure}\hfill
  \begin{subfigure}{.4\textwidth}
    \includegraphics[width=\linewidth]{assets/f1c.png}
  \end{subfigure}
  \caption{
    (a) $T_{1}$ optimized gates (red) and analytic gates (blue).
    (b) $T_{1}$ interpolation function used in optimization. Markers
    denote the time-averaged, absolute amplitude of each gate.
    (c) Lindblad master equation simulation with $T_{1}$ dissipation
    for successive gate applications. The reported gate error is computed after
    each gate appliaction.
  }
\end{figure*}

%% S4
\section{Longitudinal Relaxation Awareness}
(Strategy) We seek to minimize the probability
that the qubit decays as a result of longitudinal
relaxation. To this end we add the
longitudinal relaxation probability
to the augmented state vector
\begin{equation}
  P_{1}(t) = \int_{0}^{t} T_{1}^{-1}(a(t^{\prime})) dt^{\prime}
\end{equation}
Setting the target
probability to zero results in
a quadratic cost at each knot point
${\lvert P_{1}(t_{k}) \rvert}^{2}$.
$T_{1}(a_{k})$ is obtained at each knot point by evaluating
a spline fit to
experimental data of the form
$\{(a, T_{1})\}$.
Calculating $T_{1}$ directly from theoretical
considerations requires many high-rank
eigendecompositions, which
is computationally expensive. Additionally,
$T_{1}$ values are known to fluctuate greatly
with laboratory temperatures \cite{klimov2018fluctuations}.
Interpolating $T_{1}$ from experimental data increases
the fridge truth of the simulation.

@@

The optimizations for the $Y/2$ and $Z/2$ gates converge on similar solutions, a periodic
waveform with amplitude $\sim 0.2 \textrm{GHz}$, see Figure 1.
Both solutions extend their gate times
beyond their analytic counterpart, trading longer gate times for access
to higher amplitudes and therefore higher $T_{1}$ times. Both gates reduce
their probability of longitudinal relaxation per gate by a factor of $\sim 5$ over
their analytic counterparts, a similar scaling for their single gate error improvements
$\sim 3 \cdot 10^{-5} \rightarrow \sim 7 \cdot 10^{-6}$.
We define gate error as the infidelity of the final state and the target state
averaged over multiple pseudo-randomly generated initial states.
Furthermore, they perform simlarly in
the concatenated gate application comparision, maintaining a total gate error under $10^{-2}$
over $10^{3}$ gate applications. 
The $T_{1}$ optimized $X/2$ gate achieves a single gate error...

@@

We consider two numerical strategies for achieving robustness
to parameter deviations. The first we call the
derivative method. We draw on the intuition that
the sensitivity of the state evolution to a parameter
$\lambda$ is encoded in the $l$\textsuperscript{th}-order
derivative of the state with respect to that parameter
$\ket{\partial_{\lambda}^{l} \psi}$. The $m$\textsuperscript{th}-order
derivative method minimizes the norm of the first $m$
state derivatives with a quadratic cost at each knot point
${\lvert \braket{\partial^{l}_{\lambda} \psi_{k} | \partial^{l}_{\lambda}
    \psi_{k}} \rvert}^{2}$, $l \in \{1, \dots, m\}$.

The state derivatives could be obtained with backwards mode differentiation.
Naive automatic differentiation would compute
the state derivative at all $1, \dots, k - 1$ knot points
to obtain the state derivative at knot point $k$.
For $d$ sensitivity parameters and $N$ knot points
this requires $O(d N^2)$ matrix multiplications. Instead,
we forward propagate the state derivatives in the
augmented state vector under coupled equations of motion, resulting
in $O(d N)$ matrix multiplications.
For example, the equations of motion for the 1\textsuperscript{st}-order derivative
method are
\begin{align}
  i \hbar \frac{d}{dt} \ket{\psi} &= H \ket{\psi}\\
  i \hbar \frac{d}{dt} \ket{\partial_{\lambda}\psi} &=
  H \ket{\partial_{\lambda} \psi} +
  (\partial_{\lambda} H) \ket{\psi}
\end{align}
Matrix exponential schemes that account for the non-linear
term may be used to efficiently integrate the coupled dynamics
\cite{berland2005solving, einkemmer2017performance}.

The second scheme we analyze is the sampling method,
which has been proposed previously for
QOC \cite{reinhold2019controlling}.
We propagate additional states in the augmented state vector that follow deviant dynamics
and penalize their differences from the target state.
The states evolve under the deviant
Hamiltonian where the parameter of interest is replaced by
a deviant value $\lambda \gets \lambda + \delta \lambda$.
We propagate the additional states $\ket{\psi^{\pm}}$ with
deviant parameters $\lambda^{\pm} = \lambda \pm \sigma_{\lambda}$.
The differences between the final states and the target
state are penalized, resulting in quadratic costs at
each knot point ${\lvert \braket{\tilde{\psi}^{\pm}_{k} | \tilde{\psi}^{\pm}_{k}}
  \rvert}^{2}, \ket{\tilde{\psi}^{\pm}_{k}} = \ket{\psi^{\pm}_{k}} - \ket{\psi_{f}}$.
We sample two copies ($\ket{\psi^{\pm}}$) of four initial states with this method,
resulting in eight additional states added to the augmented state vector.
The four initial states are chosen so
that their outer products span the operators on the
Hilbert space $\{\ket{0}, \ket{1}, (\ket{0} + i\ket{1}) / \sqrt{2},
  (\ket{0} - \ket{1}) / \sqrt{2}\}$ \cite{chow2009randomized}. We observe
that sampling only the basis states $\{\ket{0}$, $\ket{1}\}$
is insufficient because, as in the case of a $Z$ rotation, their
trajectories are insensitive to some parameter deviations and therefore
they are undesirable samples for those operations.

Additionaly, we consider the unscented sampling method,
which is well-studied in the robust control iterature
\cite{julier2004unscented, lee2013sigma, manchester2016derivative}.
The strategy is to
propagate an ensemble of states (sigma points) which represent a distribution
over every element of $\ket{\psi}$. The distribution
models the uncertainty in each state element arising from the
parameter deviation. The ensemble consists of $2 (n + d)$ states
where $n$ is the size of the state vector and $d$ is the number of
deviant parameters. Each state in the ensemble is propagated to the next
knot point using separate deviant dynamics. Then, the mean and covariance of
the propagated ensemble is calculated. New states are sampled from the distribution
given by the calculated statistics for propagation to the next knot point.
This resampling procedure, the unscented transformation, accurately propagates
first and second moments of the distribution and ensures the points
lie on the $\sqrt{n}$\textsuperscript{th}
covariance contour at each knot point. A detailed update procedure is given
in the appendix. The initial state of all $2 (n + d)$ samples is
$(\ket{0} + i\ket{1} / \sqrt{2})$ with added noise. Repeating this procedure for the other
three operator basis states would provide the optimizer with more information,
but come at a computational cost which far exceeds that of the other methods,
resulting in $O(n^2)$ additional states in the augmented state vector.

@@

We compare the robust methods to an analytically derived $Z/2$ pulse which corresponds to
idling at the flux frustration point $a = 0$. The analytic pulse
is at the device's speed limit for a $Z/2$ gate $t_{Z/2} = 1 / 4 f_{q}$ and
is simple to derive. It is linearly sensitive to qubit frequency detunings,
the erroneous rotation given by $2 \pi t_{Z/2} \delta f_{q}$. At a one-percent
qubit frequency detuning, it achieves a gate error of $\sim 4.5 \cdot 10^{-5}$, sufficient
for quantum error correction.

@@

The derivative methods also converge on qualitatively similar solutions that
use fast traingle pulses at the boundaries and balance time
on either side of the flux-frustration point symmetrically at low amplitudes.
They achieve super-linear scaling in their gate error as
a function of the gate duration which approaches zero at four multiples
of $t_{Z/2}$. We observe that on a range of problems at small parameter detunings,
the 1\textsuperscript{st}-order derivative method tends to outperform the 2\textsuperscript{nd}-order
derivative method. We believe this is due to the low contribution of second-order
terms to the gate error in this deviation regime, see Table 2.

>>>>

In addition to constructing gates
that obey experimental constraints and have high simulated fidelities,
we also desire to make the gates robust to noise that affects the experimental
gate fidelity. Decoherence of the quantum state due to external noise
is typically modeled by two phenomena: longitudinal relaxation and pure dephasing.
They are modeled using their $1/e$ decay times $T_{1}$ and $T_{\phi}$ respectively
(see Appendix).
The main contributions to longitudinal relaxation in our
device are dielectric loss in the capacitor, resistive loss in the inductor,
and Purcell loss. The main contributions to pure dephasing in our
device are $1/f$ flux noise and decay via charge and flux coupling
to the control lines.

Dissipation to the thermal bath via longitudinal
relaxation is an irreversible process
that results in information loss.
Converesely, pure dephasing is a reversible process.
There is a tradeoff between the two decoherence processes. In the case of white
noise the sum of the noise weights $W_{1}$ and $W_{\phi}$
is constant \cite{huang2020engineering}.
Our device achieves its best pure dephasing
protection at the flux frustration point
$T_{2e}(a = 0) \sim 300 \mu\textrm{s}$
where the qubit is first-order insensitive to changes in flux.
It becomes more succeptable to pure dephasing as the flux is tuned away from the flux
frustration point $|a| > 0$. Conversely, longitudinal relaxation is at a minimum
at the flux frustration point $T_{1}(a = 0) = 0.315$ms,
and increases away from the flux frustration point
$T_{1}(a = 0.34 \textrm{GHz}) = 4.3$ms. Given the nature
of the decay processes and the tradeoff, we choose
to maximize the longitudinal relaxation time directly
via the flux amplitude and employ robust control techniques to mitigate
pure dephasing.

%% F1
\begin{figure*}[ht]
  \begin{subfigure}{.315\textwidth}
    \includegraphics[width=\linewidth]{assets/f1a.png}
  \end{subfigure}\hfill
  \begin{subfigure}{.23\textwidth}
    \includegraphics[width=\linewidth]{assets/f1b.png}
  \end{subfigure}\hfill
  \begin{subfigure}{.4\textwidth}
    \includegraphics[width=\linewidth]{assets/f1c.png}
  \end{subfigure}
  \caption{
    (a) $T_{1}$ optimized gates (red) and analytic gates (blue).
    (b) $T_{1}$ interpolation function used in optimization. Markers
    denote the time-averaged, absolute amplitude of each gate.
    (c) Lindblad master equation simulation with $T_{1}$ dissipation
    for successive gate applications. The reported gate error is computed after
    each gate appliaction. The $T_{1}$ optimized $Z/2$ and $Y/2$ gate errors are indistinguishable
    in the figure.
  }
\end{figure*}

%% S4
\section{Longitudinal Relaxation Awareness}
(Strategy) We seek to minimize the probability
that the qubit decays as a result of longitudinal
relaxation. To this end we add the
longitudinal relaxation probability
to the augmented state vector
\begin{equation}
  P_{1}(t) = \int_{0}^{t} T_{1}^{-1}(a(t^{\prime})) dt^{\prime}
\end{equation}
Setting the target
probability to zero results in
a quadratic cost at each knot point
${\lvert P_{1}(t_{k}) \rvert}^{2}$.
$T_{1}(a_{k})$ is obtained at each knot point by evaluating
a spline fit to
experimental data of the form
$\{(a, T_{1})\}$.
Calculating $T_{1}$ directly from theoretical
considerations requires many high-rank
eigendecompositions, which
is computationally expensive. Additionally,
$T_{1}$ values are known to fluctuate greatly
with laboratory temperatures \cite{klimov2018fluctuations}.
Interpolating $T_{1}$ from experimental data increases
the fridge truth of the simulation. Furthermore, this scheme for mitigating
longitudinal relaxation requires one
additional real number be added to the augmented state vector, avoiding
the $O(n^{2})$ increase in computational and memory complexities
associated with propagating a density matrix under the Lindblad
master equation. Here $n$ is the size of the state vector.

@@

The optimizations for the $Y/2$ and $Z/2$ gates converge on similar solutions, a periodic
waveform with amplitude $\sim 0.2 \textrm{GHz}$, see Figure 1.
Both solutions extend their gate times
beyond their analytic counterpart, trading longer gate times for access
to higher amplitudes and therefore higher $T_{1}$ times. Both gates reduce
their single gate probability of longitudinal relaxation by a factor of $5$ over
their analytic counterparts, scaling similarly to their single gate error reduction
$3 \cdot 10^{-5} \rightarrow 7 \cdot 10^{-6}$, see Appendix.
The gate error reported in this text is the infidelity of the evolved state and the target state
averaged over multiple pseudo-randomly generated initial states.
Furthermore, they perform simlarly in
the concatenated gate application comparision, maintaining total gate errors under $10^{-2}$
over $10^{3}$ gate applications. Their single gate errors are sufficient for
quantum error correction, requisite for fault-tolerant quantum computing,
and their low gate errors for extend computations are critical for noisy, intermediate-scale
quantum (NISQ) applications.
The $T_{1}$ optimized $X/2$ gate achieves a single gate error...

@@

We consider three numerical strategies for achieving robustness
to parameter deviations. The first we call the
derivative method. We draw on the intuition that
the sensitivity of the state evolution to a parameter
$\lambda$ is encoded in the $l$\textsuperscript{th}-order
derivative of the state with respect to that parameter
$\ket{\partial_{\lambda}^{l} \psi}$. The $m$\textsuperscript{th}-order
derivative method minimizes the norm of the first $m$
state derivatives with a quadratic cost at each knot point
${\lvert \braket{\partial^{l}_{\lambda} \psi | \partial^{l}_{\lambda} \psi}
  \rvert}^{2}$, $l \in \{1, \dots, m\}$.

The state derivatives could be obtained with backwards mode differentiation.
Naive automatic differentiation would compute
the state derivative at all $1, \dots, k - 1$ knot points
to obtain the state derivative at knot point $k$.
For $d$ sensitivity parameters and $N$ knot points
this requires $O(d N^2)$ matrix multiplications. Instead,
we forward propagate the state derivatives in the
augmented state vector under coupled equations of motion, resulting
in $O(d N)$ matrix multiplications.
For example, the equations of motion for the 1\textsuperscript{st}-order derivative
method are
\begin{align}
  i \hbar \frac{d}{dt} \ket{\psi} &= H \ket{\psi}\\
  i \hbar \frac{d}{dt} \ket{\partial_{\lambda}\psi} &=
  H \ket{\partial_{\lambda} \psi} +
  (\partial_{\lambda} H) \ket{\psi}
\end{align}
Exponential integrators that account for the non-linear
term may be used to efficiently integrate the coupled dynamics
\cite{berland2005solving, einkemmer2017performance}.

The second scheme we analyze is the sampling method,
which has been proposed previously for
QOC \cite{reinhold2019controlling}.
We propagate additional states in the augmented state vector that follow deviant dynamics
and penalize their differences from the target state.
The states evolve under the deviant
Hamiltonian where the parameter of interest is replaced by
a deviant value $\lambda \gets \lambda + \delta \lambda$.
We propagate the additional states $\ket{\psi^{\pm}}$ with
deviant parameters $\lambda^{\pm} = \lambda \pm \sigma_{\lambda}$.
The infidelities of the samples and the target
state are penalized at each knot point
$1 - {\lvert \braket{\psi^{\pm}_{k} | \psi_{f}} \rvert}^{2}$.
We sample two trajectories ($\ket{\psi^{\pm}}$) for each of four initial states,
resulting in eight additional states added to the augmented state vector.
The four initial states are chosen so
that their outer products span the operators on the
Hilbert space $\{\ket{0}, \ket{1}, (\ket{0} + i\ket{1}) / \sqrt{2},
  (\ket{0} - \ket{1}) / \sqrt{2}\}$ \cite{chow2009randomized}. We observe
that sampling only the basis states $\{\ket{0}$, $\ket{1}\}$
is insufficient because, as in the case of a $Z$ rotation, their
trajectories are insensitive to some parameter deviations, making
them undesirable samples for those operations.

Additionaly, we consider the unscented sampling method,
which is well-studied in the robust control iterature
\cite{julier2004unscented, lee2013sigma, manchester2016derivative}.
The strategy is to
propagate an ensemble of states (sigma points) which represent a distribution
over every element of $\ket{\psi}$. The distribution
models the uncertainty in each state element arising from the
parameter deviation. The ensemble consists of $2 (n + d)$ states
where $n$ is the size of the state vector and $d$ is the number of
deviant parameters. Each state in the ensemble is propagated to the next
knot point using separate deviant dynamics. Then, the mean and covariance of
the propagated ensemble is calculated. New states are sampled from the distribution
given by the calculated statistics for propagation to the next knot point.
This resampling procedure, the unscented transformation, accurately propagates
first and second moments of the distribution and ensures the points
lie on the $\sqrt{n}$\textsuperscript{th}
covariance contour at each knot point. A detailed update procedure is given
in the appendix. The initial state of all $2 (n + d)$ samples is
$(\ket{0} + i\ket{1} / \sqrt{2})$ with added noise. Repeating this procedure for the other
three operator basis states would provide the optimizer with more information,
but come at a computational cost which far exceeds that of the other methods,
resulting in $O(n^2)$ additional states in the augmented state vector. The
costs at each knot point are the infidelities
between the samples and the target state, as in the
basic sampling method.

@@

We compare the robust methods to an analytically derived $Z/2$ pulse which corresponds to
idling at the flux frustration point $a = 0$. The analytic pulse
is at the device's speed limit for a $Z/2$ gate $t_{Z/2} = 1 / 4 f_{q}$ and
is simple to derive. It is linearly sensitive to qubit frequency detunings,
the erroneous rotation given by $2 \pi t_{Z/2} \delta f_{q}$. At a one-percent
qubit frequency detuning, it achieves a gate error $\sim 4.5 \cdot 10^{-5}$,
sufficient for quantum error correction.

@@

The derivative methods also converge on qualitatively similar solutions that
use fast traingle pulses at the boundaries and balance time
on either side of the flux-frustration point symmetrically at low amplitudes.
They achieve super-linear scaling in their gate error as
a function of the gate duration which approaches zero at the larmor
period $1 / f_{q}$. We observe that on a range of problems at small parameter detunings,
the 1\textsuperscript{st}-order derivative method tends to outperform the 2\textsuperscript{nd}-order
derivative method. We believe this is due to the low contribution of second-order
terms to the gate error in this deviation regime, see Table 2.
